Compositional equivalences based on Open pNets

### 2021-10-01 2016-Weak-FH-Bisim/WeakBisim.tex
a few corrections in intro

In the nineties, several 
works extended the basic behavioural models based on labelled
transition systems to address value-passing or parameterised systems, using
various symbolic encodings of the
transitions~\cite{deSimone85,Larsen87}. 
These works use the term \emph{parameter} to designate systems where variables defining the system have a strong influence on their behaviour: in parameterised systems, parameters can typically influence the number of processes in the system or the way they interact.
In \cite{IngolfsdottirL:2001,HennessyLin:TCS95}, Lin, Ingolfsdottir and Hennessy developed a full hierarchy of
bisimulation equivalences, together with a proof system, for value passing CCS,
including notions of symbolic behavioural semantics and various symbolic 
bisimulations (early and late, strong and weak, and their congruent versions). They also extended this work to models with explicit assignments \cite{Linconcur96}.
Separately J. Rathke~\cite{HennessyRathke:TCS98} defined another symbolic semantics for 
a parameterised broadcast calculus, together with strong and weak bisimulation
equivalences, and developed a symbolic model-checker based on a tableau
method for these processes. 30 years later, no
practical verification approach and no verification platform are
using this kind of approaches to provide proof methods for
value-passing processes or open process expressions.

@@

\subsubsection*{pNets and open automata}
pNets is a convenient model to model concurrent systems in a hierarchical and parameterised way. The coordination between processes is expressed as synchronisation vectors that allow the definition of complex and expressive synchronisation patterns.
Open pNets are pNets for which some elements in the hierarchy are still undefined, such undefined elements are called \emph{holes}; a hole can be \emph{filled} later by providing another pNet.
 The semantics of pNets can be expressed as a translation to a labelled transition system, but only if the pNet has no parameter and no holes. Adding parameters to a LTS is quite standard but enabling holes inside LTSs is not a well-defined notion. 
We thus define \emph{open automata} that can be seen as LTSs with parameters and holes. The transitions of open automata are much more complex than transitions of an LTS as the firing on a transition depends on parameters and actions that are symbolic. This article defines the notion of \emph{open transition} a transition that is symbolic in terms of parameters and coordinated actions.

Contrarily to pNets, open automata are not hierarchical structures, thus they are more convenient for formal reasoning but not adapted to the definition of a complex and structured system like pNets. Additionally, open transitions are expressed in terms of logics more than in terms of synchronised actions, synchronisation vectors on the contrary make it easy to express synchronisations that exist in process algebra orin specification and high-level languages for distributed systems.

@@

Consequently, $t_i^{i\in I}$ defines first $I$ the set over which the
family is indexed, and then $t_i$ the elements of the family.
For example $t^{i\in\{3\}}$ is
the mapping with a single entry $t$ at index $3$; exceptionally, for mappings with
only a few entries we use the notation $(3\mapsto t)$ instead.
Note that, sentences of the form ``there exists $t_i^{i\in I}$'' means there exists $I$ and a function that maps each element of $I$ to a term $t_i$.
%The operation  $A[j \mapsto a]$ updates the value associated to $j$ in map $A$ so that 
%it 
%now corresponds to $a$.

@@

This article also uses substitutions. Applying a substitution inside a term $t$ is denoted $t\subst{y_i\gets e_i}^{i\in I}$ and consists in replacing in parallel all the occurrences of variables $y_i$ in the term $t$ by the terms $e_i$. Note that a substitution is defined by a partial function that is applied on the variables inside a term. We let $\Post$ range over partial functions that are used as substitution and use the notation $\{y_i\gets e_i\}^{i\in I}$ to define such a partial function\footnote{Note that the fact that substitutions are defined from partial function  implies that each $y_i$ is different.}. These partial functions are sometimes called \emph{substitution functions} in the following.
 $\subst{\Post}$ is thus the operation that applies,  in a parallel manner,  the substitution defined by the partial function $\Post$. $\otimes$ is a composition operator on these partial functions, such that for any term $t$ we have: $t\subst{\Post\shortotimes\Post'} = (t\subst{\Post'})\subst{\Post}$.
This property must also be valid when the substitution does not operate on all variables.
We thus define a composition operation as follows: 
%\TODO{version d eric reprise et corrigee}
%$(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$
%\ERIC{J'ai un probleme ici... quand $Post'$ agit sur des vars de $t$ qui ne sont pas substitutees par $Post$, alors les 2 defs de $\shortotimes$ ci-dessus ne sont pas equivalentes:}\\
%$F(x,z)(\subst{x\gets z}\shortotimes\subst{z\gets 1})=F(x,z)(\subst{x\gets 1})=F(1,z)$\\
%\ERIC{Alors que}\\
%$(F(x,z)\subst{x\gets z})\subst{z\gets 1})=F(z,z)\subst{z\gets 1}=F(1,1)$
%
%\ERIC{Une definition plus correcte serait peut-etre:}\\

@@

A pNet $Q$ is \emph{closed} if it has no hole: $\Holes(Q)=\emptyset$; else it
is said to be \emph{open}.
Sort comes naturally with a compatibility relation that is similar to a type-compatibility check. We simply say that two sorts are compatible if they consist of the same actions with the same arity. In practice, it is sufficient to check the equality of the two sets of pairs $(\alpha,n)$ of the two sorts.
\end{definition}

>>>>

In the nineties, several 
works extended the basic behavioural models based on labelled
transition systems to address value-passing or parameterised systems, using
various symbolic encodings of the
transitions~\cite{deSimone85,Larsen87}. 
These works use the term \emph{parameter} to designate systems where variables that have a strong influence the system structure and behaviour. In parameterised systems, parameters can typically be the number of processes in the system or the way they interact.
In \cite{IngolfsdottirL:2001,HennessyLin:TCS95}, Lin, Ingolfsdottir and Hennessy developed a full hierarchy of
bisimulation equivalences, together with a proof system, for value passing CCS,
including notions of symbolic behavioural semantics and various symbolic 
bisimulations (early and late, strong and weak, and their congruent versions). They also extended this work to models with explicit assignments \cite{Linconcur96}.
Separately J. Rathke~\cite{HennessyRathke:TCS98} defined another symbolic semantics for 
a parameterised broadcast calculus, together with strong and weak bisimulation
equivalences, and developed a symbolic model-checker based on a tableau
method for these processes. 30 years later, no
practical verification approach and no verification platform are
using this kind of approaches to provide proof methods for
value-passing processes or open process expressions.

@@

\subsubsection*{pNets and open automata}
pNets is a convenient model to model concurrent systems in a hierarchical and parameterised way. The coordination between processes is expressed as synchronisation vectors that allow the definition of complex and expressive synchronisation patterns.
Open pNets are pNets for which some elements in the hierarchy are still undefined, such undefined elements are called \emph{holes}; a hole can be \emph{filled} later by providing another pNet.
 The semantics of pNets can be expressed as a translation to a labelled transition system, but only if the pNet has no parameter and no hole. Adding parameters to a LTS is quite standard but enabling holes inside LTSs is not a well-defined notion. 
We thus define \emph{open automata} that can be seen as LTSs with parameters and holes. The transitions of open automata are much more complex than transitions of an LTS as the firing on a transition depends on parameters and actions that are symbolic. This article defines the notion of \emph{open transition} a transition that is symbolic in terms of parameters and coordinated actions.

Contrarily to pNets, open automata are not hierarchical structures, thus they are more convenient for formal reasoning but not adapted to the definition of a complex and structured system like pNets. Additionally, open transitions are expressed in terms of logics more than in terms of synchronised actions, synchronisation vectors on the contrary make it easy to express synchronisations that exist in process algebra or in specification and high-level languages for distributed systems.

@@

Consequently, $t_i^{i\in I}$ defines first $I$ the set over which the
family is indexed, and then $t_i$ the elements of the family.
For example $t^{i\in\{3\}}$ is
the mapping with a single entry $t$ at index $3$; exceptionally, for mappings with
only a few entries we use the notation $(3\mapsto t)$ instead.
In this article, sentences of the form ``there exists $t_i^{i\in I}$'' means there exists $I$ and a function that maps each element of $I$ to a term $t_i$.
%The operation  $A[j \mapsto a]$ updates the value associated to $j$ in map $A$ so that 
%it 
%now corresponds to $a$.

@@

This article also uses substitutions. Applying a substitution inside a term $t$ is denoted $t\subst{y_i\gets e_i}^{i\in I}$ and consists in replacing in parallel all the occurrences of variables $y_i$ in the term $t$ by the terms $e_i$. Note that a substitution is defined by a partial function that is applied on the variables inside a term. We let $\Post$ range over partial functions that are used as substitution and use the notation $\{y_i\gets e_i\}^{i\in I}$ to define such a partial function\footnote{When using this notation, we suppose, without loss of generality that each $y_i$ is different.}. These partial functions are sometimes called \emph{substitution functions} in the following.
 $\subst{\Post}$ is thus the operation that applies,  in a parallel manner,  the substitution defined by the partial function $\Post$. $\otimes$ is a composition operator on these partial functions, such that for any term $t$ we have: $t\subst{\Post\shortotimes\Post'} = (t\subst{\Post'})\subst{\Post}$.
This property must also be valid when the substitution does not operate on all variables.
We thus define a composition operation as follows: 
%\TODO{version d eric reprise et corrigee}
%$(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$
%\ERIC{J'ai un probleme ici... quand $Post'$ agit sur des vars de $t$ qui ne sont pas substitutees par $Post$, alors les 2 defs de $\shortotimes$ ci-dessus ne sont pas equivalentes:}\\
%$F(x,z)(\subst{x\gets z}\shortotimes\subst{z\gets 1})=F(x,z)(\subst{x\gets 1})=F(1,z)$\\
%\ERIC{Alors que}\\
%$(F(x,z)\subst{x\gets z})\subst{z\gets 1})=F(z,z)\subst{z\gets 1}=F(1,1)$
%
%\ERIC{Une definition plus correcte serait peut-etre:}\\

@@

A pNet $Q$ is \emph{closed} if it has no hole: $\Holes(Q)=\emptyset$; else it
is said to be \emph{open}.
Sort comes naturally with a compatibility relation that is similar to a type-compatibility check. We simply say that two sorts are compatible if they consist of the same actions with the same arity. In practice, it is sufficient to check the equality of the two sets of pairs $(\alpha,n)$ of the two sorts\footnote{A more complex compatibility relation could be defined, but this is out of the scope of this article.}.
\end{definition}
